New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-13 jso.qmd
modified results section and added papers code to supplementary

# Simulation study {#sec-sim-deets}

@@

The results from the first simulation described in @sec-sim-deets are
analysed based on the final projections across two optimisers (JSO and
CRS) and the success rate across JSO hyper-parameters. In the second
simulation, smoothness and squintability are calculated across a
collection of pipe-finding and sine-wave finding problems to construct
the relationship between success rate, JSO hyper-parameters, and index
properties.

The final projections found by the two optimisers (JSO and CRS) are
presented in @fig-proj, broken down by 10th quantile, faceted by the
data dimensions. In the 6-dimensional data scenario, JSO consistently
identifies a clear pipe shape. The CRS also finds the pipe shape but
with a wide rim, suggesting a further polish search may be required.
With increasing dimensions, JSO may not always identify the pipe shape
due to random sampling, but it still finds the pipe shape in over 50% of
cases. Compared to CRS, JSO achieves higher index values and clearer
pipe shapes across all quantiles in data of 8, 10, 12 dimensions,
suggesting its advantage in exploring high-dimensional spaces.

The success rate calculated at each hyper-parameter combination (number
of jellyfish and the maximum number of tries) is presented in
@fig-proportion. The uncertainty is quantified through 500 bootstrap replicates for each case. As the number of jellyfish and maximum tries increase, the success rate also increases. For simpler problems (6 dimensions),
small parameter values (20 jellyfishes and a maximum of 50 tries) can
already achieve a high success rate. However, larger parameter values
(i.e. 100 jellyfishes and a maximum of 100 tries) are necessary for
higher-dimensional problems (8, 10, and 12 dimensions). Increasing both
parameters enhances the performance of JSO, but it also extends the
computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. @tbl-smoothness-squintability presents the
parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function ($\theta_1$ to $\theta_4$) for each case considered in the
simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation (@eq-squintability-parametric)
as the squintability metric. @tbl-mod-output presents the results of
fitting a generalised linear model with a binomial family and a logit
link function. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
`r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`% and
`r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success.

@@

```{r}
#| results: "asis"
stargazer::stargazer(
  mod1, header = FALSE, label = "tbl-mod-output",
  table.placement = "H", 
  covariate.labels = c("Smoothness", "Squintability", "Dimension", "Long runtime", "Number of jellyfish", "Maximum number of tries"),
  title = "Model estimates of proportion of jellyfish success on index properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function.The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with JSO success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success. ")
```

@@

The supplementary materials include: 1) description of all the index functions used in the simulations and 2) the script that complements the Practicalities section for using JSO in a PPGT and calculating smoothness and squintability.

>>>>

# Assessing optimisers: a simulation study {#sec-sim-deets}

@@

## Performance of JSO relative to CRS

Performance is assessed based on the final projections across two optimisers (JSO and CRS) and the success rate across JSO hyper-parameters. The final projections found by the two optimisers (JSO and CRS) are
presented in @fig-proj, broken down by 10th quantile, faceted by the
data dimensions. In the 6-dimensional data scenario, JSO consistently
identifies a clear pipe shape. The CRS also finds the pipe shape but
with a wide rim, suggesting a further polish search may be required.
With increasing dimensions, JSO may not always identify the pipe shape
due to random sampling, but it still finds the pipe shape in over 50% of
cases. Compared to CRS, JSO achieves higher index values and clearer
pipe shapes across all quantiles in data of 8, 10, 12 dimensions,
suggesting its advantage in exploring high-dimensional spaces.

## Factors affecting JSO success rate: index properties and jellyfish hyper-parameters

Smoothness and squintability are calculated across a collection of pipe-finding and sine-wave finding problems to construct the relationship between success rate, JSO hyper-parameters, and index properties. The success rate calculated at each hyper-parameter combination (number of jellyfish and the maximum number of tries) is presented in @fig-proportion. The uncertainty is quantified through 500 bootstrap replicates for each case. As the number of jellyfish and maximum tries increase, the success rate also increases. For simpler problems (6 dimensions), small parameter values (20 jellyfishes and a maximum of 50 tries) can already achieve a high success rate. However, larger parameter values
(i.e. 100 jellyfishes and a maximum of 100 tries) are necessary for
higher-dimensional problems (8, 10, and 12 dimensions). Increasing both
parameters enhances the performance of JSO, but it also extends the
computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

The index properties, smoothness and squintability, numerically characterise the complexity of projection pursuit index, and hence potential optimisation problems. @tbl-smoothness-squintability presents the parameters for calculating both metrics estimated from the Gaussian process (variance, range, smooth, and nugget) and the scaled logistic function ($\theta_1$ to $\theta_4$) for each case considered in the simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation (@eq-squintability-parametric)
as the squintability metric.

@tbl-mod-output presents the results from fitting a logistic regression model using proportion of success as the response and smoothness, squintability, dimension, running time, number of jellyfish and maximum number of tries, as predictors. Long runtime is a binary variable, where greater than 20 seconds is labelled 1. The fit suggests that JSO success rate is only affected by squintability, dimension and number of jellies. Interestingly, the smoothness of the index is not a problem for the JSO - this is promising! Running the optimisation longer and allowing more tries also does not affect success rate.  An increase in squintability of one unit increases the success rate by `r round(exp(broom::tidy(mod1)$estimate[3]) * 100, 0)`%. As dimension increases by one the success rate halves. Increasing the number of jellies by 10, increases the succes rate by `r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 0)`%.

@@

```{r}
#| label: tbl-mod-output
#| tbl-cap: "Jellyfish success rate relative to index properties and jellyfish hyper-parameters. This is the summary from a logistic regression fit to smoothness, squintability, dimension, running time, number of jellyfish and maximum number of tries. Interestingly, squintability and dimension strongly affect jellyfish optimisation success. The number of jellies marginally affects success, but index smoothness, running longer and increasing the number of tries do not."
#| fig-pos: t
library(gtsummary)
set_gtsummary_theme(theme_gtsummary_journal("jama"))
mod1 |> tbl_regression(exponentiate=TRUE, 
                       pvalue_fun = label_style_pvalue(digits = 3),
                       label = list(smoothness = "Smoothness", 
                                    squintability = "Squintability",
                                    d = "Dimension", 
                                    long_time = "Long runtime", 
                                    n_jellies = "Number of jellyfish", 
                                    max_tries = "Maximum number of tries")) |>
  bold_p(t = 0.1)
```

@@

The supplementary materials available at <https://github.com/huizezhang-sherry/paper-jso> include: 1) details of the indexes used in the simulation study, 2) the script to get started with using JSO in a PPGT and calculating smoothness and squintability, as explained in @sec-discussion, and the full code to reproduce the plots and summaries in this paper.

### 2024-09-13 jso.tex
modified results section and added papers code to supplementary

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{orcidlink}
\usepackage{algorithm}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\theoremstyle{plain}
\newtheorem{assumption}{\protect\assumptionname}
\newtheorem{claim}{\protect\claimname}
\newtheorem{condition}{\protect\conditionname}
\newtheorem{conjecture}{\protect\conjecturename}
\newtheorem{cor}{\protect\corollaryname}
\newtheorem{defn}{\protect\definitionname}
\newtheorem{example}{\protect\examplename}
\newtheorem{lem}{\protect\lemmaname}
\newtheorem{notation}{\protect\notationname}
\newtheorem{problem}{\protect\problemname}
\newtheorem{prop}{\protect\propositionname}
\newtheorem{rem}{\protect\remarkname}
\newtheorem{thm}{\protect\theoremname}
\providecommand{\assumptionname}{Assumption}
\providecommand{\claimname}{Claim}
\providecommand{\conditionname}{Condition}
\providecommand{\conjecturename}{Conjecture}
\providecommand{\corollaryname}{Corollary}
\providecommand{\definitionname}{Definition}
\providecommand{\examplename}{Example}
\providecommand{\lemmaname}{Lemma}
\providecommand{\notationname}{Notation}
\providecommand{\problemname}{Problem}
\providecommand{\propositionname}{Proposition}
\providecommand{\remarkname}{Remark}
\providecommand{\theoremname}{Theorem}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, leftrule=.75mm, colframe=quarto-callout-note-color-frame, title={Algorithm: Jellyfish Optimizer Pseudo Code}, rightrule=.15mm, coltitle=black, colback=white, bottomrule=.15mm, bottomtitle=1mm, toptitle=1mm, left=2mm, titlerule=0mm, arc=.35mm, toprule=.15mm, opacitybacktitle=0.6]

@@

\section{Simulation study}\label{sec-sim-deets}

@@

The results from the first simulation described in
Section~\ref{sec-sim-deets} are analysed based on the final projections
across two optimisers (JSO and CRS) and the success rate across JSO
hyper-parameters. In the second simulation, smoothness and squintability
are calculated across a collection of pipe-finding and sine-wave finding
problems to construct the relationship between success rate, JSO
hyper-parameters, and index properties.

The final projections found by the two optimisers (JSO and CRS) are
presented in Figure~\ref{fig-proj}, broken down by 10th quantile,
faceted by the data dimensions. In the 6-dimensional data scenario, JSO
consistently identifies a clear pipe shape. The CRS also finds the pipe
shape but with a wide rim, suggesting a further polish search may be
required. With increasing dimensions, JSO may not always identify the
pipe shape due to random sampling, but it still finds the pipe shape in
over 50\% of cases. Compared to CRS, JSO achieves higher index values
and clearer pipe shapes across all quantiles in data of 8, 10, 12
dimensions, suggesting its advantage in exploring high-dimensional
spaces.

The success rate calculated at each hyper-parameter combination (number
of jellyfish and the maximum number of tries) is presented in
Figure~\ref{fig-proportion}. The uncertainty is quantified through 500
bootstrap replicates for each case. As the number of jellyfish and
maximum tries increase, the success rate also increases. For simpler
problems (6 dimensions), small parameter values (20 jellyfishes and a
maximum of 50 tries) can already achieve a high success rate. However,
larger parameter values (i.e.~100 jellyfishes and a maximum of 100
tries) are necessary for higher-dimensional problems (8, 10, and 12
dimensions). Increasing both parameters enhances the performance of JSO,
but it also extends the computational time required for the
optimisation, which can be computationally intensive when evaluating the
index function (such as scagnostic indexes) multiple times across
numerous iterations.

@@

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. Table~\ref{tbl-smoothness-squintability} presents
the parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function (\(\theta_1\) to \(\theta_4\)) for each case considered in the
simulation. The column ``smooth'' is used as the smoothness metric and
the column ``squint'' is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.
Table~\ref{tbl-mod-output} presents the results of fitting a generalised
linear model with a binomial family and a logit link function. The model
suggests that JSO success rate is positively associated with the two
hyper-parameters, as well as with the index properties: smoothness and
squintability. Specifically, using 10 more jellyfish and 10 more tries
increases the odd ratio of success by 24.11\% and 11.93\%, respectively.
However, being flagged with long runtime and an increase of data
dimension reduce the success rate by 41.72\% and 53.36\%, respectively.
The variable \texttt{squintability} and \texttt{dimension} are
significant, suggesting their importance relative to JSO
hyper-parameters in the optimisation success.

@@

\caption{\label{tbl-mod-output}}

\begin{tabular}{@{\extracolsep{5pt}}lc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{1}{c}{\textit{Dependent variable:}} \\ 
\cline{2-2} 
\\[-1.8ex] & proportion \\ 
\hline \\[-1.8ex] 
 Smoothness & 1.195 \\ 
  & (1.915) \\ 
  & \\ 
 Squintability & 2.060$^{***}$ \\ 
  & (0.684) \\ 
  & \\ 
 Dimension & $-$0.628$^{**}$ \\ 
  & (0.255) \\ 
  & \\ 
 Long runtime & $-$0.874 \\ 
  & (1.286) \\ 
  & \\ 
 Number of jellyfish & 0.216$^{*}$ \\ 
  & (0.127) \\ 
  & \\ 
 Maximum number of tries & 0.113 \\ 
  & (0.150) \\ 
  & \\ 
 Constant & $-$4.523 \\ 
  & (5.332) \\ 
  & \\ 
\hline \\[-1.8ex] 
Observations & 52 \\ 
Log Likelihood & $-$18.224 \\ 
Akaike Inf. Crit. & 50.447 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular}

@@

The supplementary materials include: 1) description of all the index
functions used in the simulations and 2) the script that complements the
Practicalities section for using JSO in a PPGT and calculating
smoothness and squintability.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{orcidlink}
\usepackage{algorithm}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\theoremstyle{plain}
\newtheorem{assumption}{\protect\assumptionname}
\newtheorem{claim}{\protect\claimname}
\newtheorem{condition}{\protect\conditionname}
\newtheorem{conjecture}{\protect\conjecturename}
\newtheorem{cor}{\protect\corollaryname}
\newtheorem{defn}{\protect\definitionname}
\newtheorem{example}{\protect\examplename}
\newtheorem{lem}{\protect\lemmaname}
\newtheorem{notation}{\protect\notationname}
\newtheorem{problem}{\protect\problemname}
\newtheorem{prop}{\protect\propositionname}
\newtheorem{rem}{\protect\remarkname}
\newtheorem{thm}{\protect\theoremname}
\providecommand{\assumptionname}{Assumption}
\providecommand{\claimname}{Claim}
\providecommand{\conditionname}{Condition}
\providecommand{\conjecturename}{Conjecture}
\providecommand{\corollaryname}{Corollary}
\providecommand{\definitionname}{Definition}
\providecommand{\examplename}{Example}
\providecommand{\lemmaname}{Lemma}
\providecommand{\notationname}{Notation}
\providecommand{\problemname}{Problem}
\providecommand{\propositionname}{Proposition}
\providecommand{\remarkname}{Remark}
\providecommand{\theoremname}{Theorem}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, arc=.35mm, colback=white, title={Algorithm: Jellyfish Optimizer Pseudo Code}, titlerule=0mm, coltitle=black, bottomrule=.15mm, left=2mm, breakable, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, bottomtitle=1mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm]

@@

\section{Assessing optimisers: a simulation study}\label{sec-sim-deets}

@@

\subsection{Performance of JSO relative to
CRS}\label{performance-of-jso-relative-to-crs}

Performance is assessed based on the final projections across two
optimisers (JSO and CRS) and the success rate across JSO
hyper-parameters. The final projections found by the two optimisers (JSO
and CRS) are presented in Figure~\ref{fig-proj}, broken down by 10th
quantile, faceted by the data dimensions. In the 6-dimensional data
scenario, JSO consistently identifies a clear pipe shape. The CRS also
finds the pipe shape but with a wide rim, suggesting a further polish
search may be required. With increasing dimensions, JSO may not always
identify the pipe shape due to random sampling, but it still finds the
pipe shape in over 50\% of cases. Compared to CRS, JSO achieves higher
index values and clearer pipe shapes across all quantiles in data of 8,
10, 12 dimensions, suggesting its advantage in exploring
high-dimensional spaces.

\subsection{Factors affecting JSO success rate: index properties and
jellyfish
hyper-parameters}\label{factors-affecting-jso-success-rate-index-properties-and-jellyfish-hyper-parameters}

Smoothness and squintability are calculated across a collection of
pipe-finding and sine-wave finding problems to construct the
relationship between success rate, JSO hyper-parameters, and index
properties. The success rate calculated at each hyper-parameter
combination (number of jellyfish and the maximum number of tries) is
presented in Figure~\ref{fig-proportion}. The uncertainty is quantified
through 500 bootstrap replicates for each case. As the number of
jellyfish and maximum tries increase, the success rate also increases.
For simpler problems (6 dimensions), small parameter values (20
jellyfishes and a maximum of 50 tries) can already achieve a high
success rate. However, larger parameter values (i.e.~100 jellyfishes and
a maximum of 100 tries) are necessary for higher-dimensional problems
(8, 10, and 12 dimensions). Increasing both parameters enhances the
performance of JSO, but it also extends the computational time required
for the optimisation, which can be computationally intensive when
evaluating the index function (such as scagnostic indexes) multiple
times across numerous iterations.

@@

The index properties, smoothness and squintability, numerically
characterise the complexity of projection pursuit index, and hence
potential optimisation problems.
Table~\ref{tbl-smoothness-squintability} presents the parameters for
calculating both metrics estimated from the Gaussian process (variance,
range, smooth, and nugget) and the scaled logistic function
(\(\theta_1\) to \(\theta_4\)) for each case considered in the
simulation. The column ``smooth'' is used as the smoothness metric and
the column ``squint'' is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.

Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using proportion of success as the response and
smoothness, squintability, dimension, running time, number of jellyfish
and maximum number of tries, as predictors. Long runtime is a binary
variable, where greater than 20 seconds is labelled 1. The fit suggests
that JSO success rate is only affected by squintability, dimension and
number of jellies. Interestingly, the smoothness of the index is not a
problem for the JSO - this is promising! Running the optimisation longer
and allowing more tries also does not affect success rate. An increase
in squintability of one unit increases the success rate by 785\%. As
dimension increases by one the success rate halves. Increasing the
number of jellies by 10, increases the succes rate by 24\%.

@@

\setlength{\LTpost}{0mm}

\begin{longtable}{lcc}

\caption{\label{tbl-mod-output}Jellyfish success rate relative to index
properties and jellyfish hyper-parameters. This is the summary from a
logistic regression fit to smoothness, squintability, dimension, running
time, number of jellyfish and maximum number of tries. Interestingly,
squintability and dimension strongly affect jellyfish optimisation
success. The number of jellies marginally affects success, but index
smoothness, running longer and increasing the number of tries do not.}

\tabularnewline

\toprule
\textbf{Characteristic} & \textbf{OR} \textbf{(95\% CI)}\textsuperscript{\textit{1}} & \textbf{p-value} \\ 
\midrule\addlinespace[2.5pt]
Smoothness & 3.30 (0.08 to 274) & 0.533 \\ 
Squintability & 7.85 (2.40 to 37.7) & \textbf{0.003} \\ 
Dimension & 0.53 (0.30 to 0.83) & \textbf{0.014} \\ 
Long runtime & 0.42 (0.03 to 5.44) & 0.497 \\ 
Number of jellyfish & 1.24 (0.98 to 1.63) & \textbf{0.089} \\ 
Maximum number of tries & 1.12 (0.84 to 1.52) & 0.452 \\ 
\bottomrule

\end{longtable}

\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}OR = Odds Ratio, CI = Confidence Interval\\
\end{minipage}

@@

The supplementary materials available at
\url{https://github.com/huizezhang-sherry/paper-jso} include: 1) details
of the indexes used in the simulation study, 2) the script to get
started with using JSO in a PPGT and calculating smoothness and
squintability, as explained in Section~\ref{sec-discussion}, and the
full code to reproduce the plots and summaries in this paper.
